Journey: A Study of Scalabilty of Memory Management in Linux

### 2017-05-05 paper.tex
edited kswapd sections

% TODO: should we mention slab allocator?
To measure memory usage, we use the procfs's reporting on current memory usage.
Where appropriate, we record the number of active
\texttt{vm\_area\_struct}s in the system. Measuring kernel memory usage and
associating memory usage with a particular process is difficult. Our methodology assumes
that our benchmarks are running in an otherwise idle system and should dominate
increases in memory usage.  We compute total memory usage by the rest of the
system as the total amount of memory less the amount of memory used by the
benchmark. We keep transparent huge pages on throughout our experiments, unless
otherwise specified.

@@

In all these experiments, the system is given 32GB of swap space total. We
repeat each experiment with hugepages enabled and disabled to see the impact of
the transparent hugepage system. The processor usage of kswapd and khugepaged is
recorded for each experiment and reported in the next section.

@@

Transparent huge pages can consume a large fraction of memory and lead to early
onset of memory pressure when the workload accesses many pages that are
separated in its address space. The overhead of swapping greatly exceeds the
performance benefits provided by huge pages. As a result, the workload
performance can fall drastically in such cases. It might be advisable to
disable transparent huge pages for such applications. The kernel can also be
cautious and limit the allocation of huge pages to processes in order to
prevent early onset of memory pressure.

@@

This experiment is designed to measure the impact of swapping on memory
operation latency. Figure \ref{fig:swap_time_cont} and Figure
\ref{fig:swap_time_rand} show the results.

@@

@@

@@

\subsection{kswapd and khugepaged (TODO: markm)}

TODO: in progress... should be ready for final product...

@@

TODO

% TODO: we need to discuss implications for large memory in the insights sections more

@@

The emergence of the internet required businesses to buy expensive machines to
keep up with demand for their services, but systems researchers did not have
access to these machines. Alameldeen et al. describe how they simulate a
multi-million dollar server using a \$2000 workstation. To do this, they go
through several rounds of scaling down, optimizing, changing, and tuning both
the benchmarks and systems they test. For example, they scale down a workload
to fit in the 1GB memory of the machine they use and increase the number of
threads to improve parallelism \cite{2kmachine}. This methodology may be
accurate, but it requires modifying the benchmarks, which is error prone and
time consuming.  Ideally, researchers can run large experimental workloads
without changes, but the overheads of current memory management systems prevent
this from being practical.

@@

%Two threads walk into a bar. Bartender Hasnain looks up and yells, ``Hey, I want
%don’t any conditions race like time last!''

The Simics simulator uses demand paging for simulated environments.
As long as the working set of
the target system fits in the host memory, performance will be tolerable.
However, in our work, we wish to study OS behavior under workloads that
actually use all available memory. Thus, the usefulness of demand paging is
greatly reduced because swapping causes disk latency to become the main
performance bottleneck. Moreover, the performance overhead of simulators like
Simics tends to be impractical because system state needs to be maintained by
the simulator \cite{simics}. We do not explore simulation further because it
can be orders of magnitude slower than native execution, making it infeasible
for studying large workloads or systems \cite{2kmachine}.

>>>>

% TODO: should we mention slab allocator?
% (MM) I don't thinks so...
To measure memory usage, we use the procfs's reporting on current memory usage.
Where appropriate, we record the number of active
\texttt{vm\_area\_struct}s in the system. Measuring kernel memory usage and
associating memory usage with a particular process is difficult. Our methodology assumes
that our benchmarks are running in an otherwise idle system and should dominate
increases in memory usage.  We compute total memory usage by the rest of the
system as the total amount of memory less the amount of memory used by the
benchmark. We keep transparent huge pages on throughout our experiments, unless
otherwise specified.

@@

In all these experiments, the system is given 32GB of swap space total. We
repeat each experiment with hugepages enabled and disabled to see the impact of
the transparent hugepage system. The processor usage of kswapd is
recorded for each experiment and reported in the next section.

@@

Transparent huge pages can consume a large fraction of memory and lead to early
onset of memory pressure for workloads with random access pattners. The overhead of swapping greatly exceeds the
performance benefits provided by huge pages. As a result, the workload
performance can fall drastically in such cases. It might be advisable to
disable transparent huge pages for such applications. The kernel can also be
cautious and limit the allocation of huge pages to processes in order to
prevent early onset of memory pressure.

@@

This experiment is designed to measure the impact of swapping on memory
operation latency.

@@

The latency of our disk leads us to expect that swapping out pages should take
roughly 4ms. However, we see that in both benchmarks, there are operations that
take dozens or hundreds of milliseconds. We used the \texttt{ftrace} kernel
profiling mechanism in Linux to find out what causes some operations to be
extremely slow.

Surprisingly, the majority of this latency comes from kswapd itself. kswapd goes
through phases of moving pages to the inactive list and reclaiming pages from
the inactive list. This repeated until the amount of free memory increases
beyond a threshold. kswapd does not do any particularly expensive operations
beyond disk I/O. Rather, it executes a bunch of cheaper operations (e.g.
checking if a page was referenced or freeing a page) millions of times in deeply
nested loops. Meanwhile, the workload continues to use up any memory reclaimed
by kswapd. Whenever kswapd is invoked by the page fault handler, it blocks the
faulting process until enough pages are reclaimed.  Moreover, in some cases,
during the \texttt{cont} workload kswapd chooses pages to reclaim much faster
than they can be written to disk.

@@

Our results also show that in some cases, kswapd induces more overhead to choose
a good page to reclaim than would be accrued by choosing the wrong page to
reclaim. Most of this overhead seems to be due to the poor asymptotic complexity
of PFRA. Moreover, the swapping subsystem should be aware of disk bandwidth and
latency, since they fundamentally limits how much memory can be reclaimed per
unit time.

Systems with larger memory may conceivably support more memory overcommitment,
so improving the performance of swapping may be highly relavent moving forward.
Additionally, if a system has orders of magnitude more pages, kswapd may spend
significantly more time scanning the active and inactive lists.

@@

Figure (TODO) shows the total number of jiffies spent running kswapd as the
number of operations performed increases. Notice that one jiffy corresponds to
10ms of runtime.

First, we find that number of jiffies is a very imperfect measure of CPU runtime
for kernel threads because it often misses executions of PFRA in the context of
page faulting processes. Due to time constraints, we were not able to rerun our
experiments using a better metric. However, these results do complement our
findings from the previous set of experiments by showing the amount of time
kswapd runs in its own context.

kswapd does not take up a significant percentage of foreground runtime compared
with the applcation. The behavior of kswapd depends heavily on the workload and
on whether transparent hugepages are enabled. Not surprisingly, workloads that
do not exhibit memory pressure do not incur kswapd overhead, as shown by the
\texttt{rand} benchmark with hugepages off.

As expected, the \texttt{rand} benchmark with hugepages on incurs the highest
total overhead since it begins swapping very early. For this benchmark we
observe an overhead of roughly 0.0002 jiffies per page memory touched after
swapping begins. Also, kswapd runs for longer periods of time with longer
intervals of sleep in between. This is most likely because the benchmark has
some probability of retouching a page, so free memory is used more slowly, but
kswapd also has to do more work to find pages that are inactive.

In contrast, the \texttt{cont} benchmark with hugepages disabled uses
computational time at roughly 0.0572 jiffies per page of memory touched.
kswapd runs more frequently and for shorter periods of time because new pages
are touched very quickly, requiring faster reclamation, but because all pages
are only touched once, kswapd has to do very little work to choose pages to
reclaim.

Notice also that the \texttt{cont} benchmark with hugepages enabled spends
surprisingly little time in kswapd. We suspect that kswapd runs once and choose
a large number of pages to swap out. Subsequently, both the workload and kswapd
wait for I/O. This seems consistent with our previous results, in which some
operations take on the order of seconds to complete, but we have not yet
verified it.

@@

The swapping system in Linux has complicated interactions with workloads and
transparent hugepages. It is not obvious how to optimize a workload for better
swapping behavior. Moreover, we see that under memory pressure, kswapd tends to
run more often during page faults than in its own context.

@@

The emergence of the internet required businesses to buy expensive machines to
keep up with demand for their services, but systems researchers did not have
access to these machines. Alameldeen et al. describe how they simulate a
multi-million dollar server using a \$2000 workstation. To do this, they go
through several rounds of scaling down, optimizing, changing, and tuning both
the benchmarks and systems they test. For example, they scale down a workload to
fit in the 1GB memory of the machine they use and increase the number of threads
to improve parallelism \cite{2kmachine}. This methodology may be accurate, but
it requires modifying the benchmarks, which is error prone and time consuming;
it can be extremely difficult to validate changes to a benchmark. Ideally,
researchers can run large experimental workloads without changes, but the overheads
of current memory management systems prevent this from being practical.

@@

Two threads walk into a bar. Bartender Hasnain looks up and yells, ``Hey, I want
don’t any conditions race like time last!''

The Simics simulator uses demand paging for simulated environments, which means
that memory is only allocated when it is used. As long as the working set of
the target system fits in the host memory, performance will be tolerable.
However, in our work, we wish to study OS behavior under workloads that
actually use all available memory. Thus, the usefulness of demand paging is
greatly reduced because swapping causes disk latency to become the main
performance bottleneck. Moreover, the performance overhead of simulators like
Simics tends to be impractical because system state needs to be maintained by
the simulator \cite{simics}. We do not explore simulation further because it
can be orders of magnitude slower than native execution, making it infeasible
for studying large workloads or systems \cite{2kmachine}. However, the Simics
simulator illustrates the usefulness of memory management techniques such as
demand paging and the potential overheads of swapping.
